---
title: "DIscBIO tutorial"
author: "Salim Ghannoum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIscBIO tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r options, echo=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=7, eval=FALSE)
```

The pipeline consists of four successive steps: data pre-processing, cellular clustering and pseudo-temporal ordering, determining differential expressed genes and identifying biomarkers.

This vignette is a reduced version of the interactive notebook available [here](https://mybinder.org/v2/gh/SystemsBiologist/PSCAN/discbio-pub?filepath=DIscBIO.ipynb)

## Loading dataset

The dataset should be formatted in a data frame where columns refer to samples and rows refer to genes.

```{r}
library(DIscBIO)
FileName <- data("valuesG1msRed")
DataSet <- get(FileName)
```

## 1. Data Pre-processing

Prior to applying data analysis methods, it is standard to pre-process the raw read counts resulted from the sequencing. The preprocessing approach depends on the existence or absence of ERCC spike-ins. In both cases, it includes normalization of read counts and gene filtering.

#### Normalization of read counts
To account for RNA composition and sequencing depth among samples (single-cells), the normalization method “median of ratios” is used. This method takes the ratio of the gene instantaneous median to the total counts for all genes in that cell (column median). The gene instantaneous median is the product of multiplying the median of the total counts across all cells (row median) with the read of the target gene in each cell. This normalization method makes it possible to compare the normalized counts for each gene equally between samples.

#### Gene filtering
The key idea in filtering genes is to appoint the genes that manifest abundant variation across samples. Filtering genes is a critical step due to its dramatic impact on the downstream analysis. In case the raw data includes ERCC spike-ins, genes will be filtered based on variability in comparison to a noise level estimated from the ERCC spike-ins according to an algorithm developed by Brennecke et al (Brennecke et al., 2013). This algorithm utilizes the dependence of technical noise on the average read count and fits a model to the ERCC spike-ins. Further gene filtering can be implemented based on gene expression.
In case the raw data does not include ERCC spike-ins, genes will be only filtered based on minimum expression in certain number of cells.

```{r}
sc <- DISCBIO(DataSet)
```

### 1.1. Filtering and normalizing the raw data that includes ERCCs

Filtering the raw data that includes ERCCs can be done by applying the “NoiseFiltering” function, which includes several parameters:

  - object: the outcome of running the DISCBIO() function.
  - percentile: A numeric value of the percentile. It is used to validate the ERCC spik-ins. Default is 0.8.
  - CV: A numeric value of the coefficient of variation. It is used to validate the ERCC spik-ins. Default is 0.5.
  - geneCol: Color of the genes that did not pass the filtration.
  - FgeneCol: Color of the genes that passt the filtration.
  - erccCol: Color of the ERCC spik-ins.
  - Val: A logical vector that allows plotting only the validated ERCC spike-ins. Default is TRUE. If Val=FALSE will
    plot all the ERCC spike-ins.
  - plot: A logical vector that allows plotting the technical noise. Default is TRUE.
  - export: A logical vector that allows writing the final gene list in excel file. Default is TRUE.
  - quiet: if `TRUE`, suppresses printed output

To normalize the raw sequencing reads the function Normalizedata() should be used, this function takes 8 parameters.
- In case the user would like just to normalize the reads without any further gene filtering the parameters minexpr and minnumber should be set to 0.
- In case the user would like just to normalize the reads and run gene filtering based on gene expression the parameters minexpr and minnumber should have values. This function will discard cells with less than mintotal transcripts. Genes that are not expressed at minexpr transcripts in at least minnumber cells are discarded.

The function Normalizedata() normalizes the count reads using the normalization method “median of ratios”

To Finalize the preprocessing the function FinalPreprocessing() should be implemented by setting the parameter "GeneFlitering" to NoiseF ( whether the dditional gene filtering step based on gene expression was implemented on not).


```{r}
sc <- NoiseFiltering(sc,percentile=0.9, CV=0.2)

####  Normalizing the reads without any further gene filtering
sc <- Normalizedata(sc, mintotal=1000, minexpr=0, minnumber=0, maxexpr=Inf, downsample=FALSE, dsn=1, rseed=17000)

sc <- FinalPreprocessing(sc,GeneFlitering="NoiseF") ### The GeneFiltering can be either "NoiseF" or"ExpF"
```

### 1.2. Filtering and normalizing the raw data that does not include ERCCs

To normalize and filter the raw data that does not include ERCCs can be done by applying the function Normalizedata() and giving  the parameters minexpr and minnumber some values. This function will discard cells with less than mintotal transcripts. Genes that are not expressed at minexpr transcripts in at least minnumber cells are discarded. Furthermore, it will normalize the count reads using the normalization method “median of ratios”.

To Finalize the preprocessing the function FinalPreprocessing() should be implemented by setting the parameter "GeneFlitering" to ExpF.

## 2. Cellular Clustering and Pseudo Time ordering

Cellular clustering is performed according to the gene expression profiles to detect cellular sub-population with unique properties. After clustering, pseudo-temporal ordering is generated to indicate the cellular differentiation degree.

## 2.1. K-means Clustering

Rare cell type Identification algorithm (RaceID) was used to cluster the pre-processed data using k-means on a similarity distance matrix, which was based on Pearson correlation and the similarity matrix was computed as “1 – Pearson correlation”. The approach of the proposed clustering, i.e., applying k-means on a similarity distance matrix using the Euclidean metric, improves cluster separation. RaceID estimates the number of clusters by finding the minimal clusters' number at the saturation level of gap statistics, which standardizes the within-cluster dispersion.

The Clustexp() functions has several parameters:
- object: the outcome of running the DISCBIO() function.
- clustnr Maximum number of clusters for the derivation of the cluster number by the saturation of mean within-cluster dispersion. Default is 20.
- bootnr A numeric value of booststrapping runs for \code{clusterboot}. Default is 50.
- metric Is the method to transform the input data to a distance object.
- Metric has to be one of the following: ["spearman","pearson","kendall","euclidean","maximum","manhattan","canberra","binary","minkowski"].
- do.gap A logical vector that allows generating the number of clusters based on the gap statistics. Default is TRUE.
- SE.method The SE.method determines the first local maximum of the gap statistics.
- The SE.method has to be one of the following:["firstSEmax","Tibs2001SEmax","globalSEmax","firstmax","globalmax"]. Default is "Tibs2001SEmax"
- SE.factor A numeric value of the fraction of the standard deviation by which the local maximum is required to differ from the neighboring points it is compared to. Default is 0.25.
- B.gap Number of bootstrap runs for the calculation of the gap statistics. Default is 50
- cln Number of clusters to be used. Default is \code{NULL} and the cluster number is inferred by the saturation criterion.
- rseed Integer number. Random seed to enforce reproducible clustering results. Default is 17000.
- quiet if `TRUE`, intermediate output is suppressed

```{r}
sc <- Clustexp(sc,cln=2,quiet=TRUE) #### K-means clustering to get three clusters
plotGap(sc) ### Plotting gap statisticssc <- Clustexp(sc, clustnr=20,bootnr=50,metric="pearson",do.gap=TRUE,SE.method="Tibs2001SEmax",SE.factor=.25,B.gap=50,cln=2,rseed=17000)
```

#### 2.1.1. Defining the Cells in the clusters generated by k-means clustering


```{r}
sc <- comptSNE(sc,rseed=15555,quiet=TRUE)
sc@cpart
```

#### 2.1.2- Evaluating the stability and consistency of the clusters

PSCAN enables the robustness assessment of the detected clusters in terms of stability and consistency using Jaccard’s similarity statistics and silhouette coefficients. Jaccard’s similarity index provides a comparison of members among clusters to evaluate the stability of the clusters with a range from 0% to 100%. The higher the percentage, the more stable the cluster is. Silhouette coefficients estimate how close each sample in one cluster is to samples in the neighboring clusters, reflecting the consistency of each cluster with a range of [-1, 1]. The higher the cluster mean coefficient, the more consistent the cluster is.


```{r}
# Silhouette of k-means clusters
# you might have to run "par(mar=c(6,2,4,2))" first so plot looks better
plotSilhouette(sc,K=2) # K is the number of clusters
```

```{r}
Jaccard(sc,Clustering="K-means", K=2, plot=TRUE) # Jaccard of k-means clusters
```

#### 2.1.3. Cluster plotting using tSNE maps

Here you visualize the K-means clusters using t-distributed stochastic neighbor embedding (tSNE), which is a non-linear dimensionality reduction method that places neighbor cells close to each other.


```{r}
############ Plotting K-means clusters
plottSNE(sc)
plotKmeansLabelstSNE(sc) # To plot the the ID of the cells in each cluster
plotSymbolstSNE(sc,types=sub("(\\_\\d+)$","", names(sc@ndata))) # To plot the the ID of the cells in each cluster
```

### Defining outlier cells based on K-means Clustering

Outlier identification is implemented using a background model based on distribution of transcript counts within a cluster. Outlier cells are detected if the probability for that cell c that a minimum number of genes Gmin of observing total counts TGmin is less than a specific threshold Pthr, as given by the red dotted line. Outlier cells in K-means clusters can be detected by running the FindOutliersKM() function.

In case the user decided to remove outlier cells, the user should set RemovingOutliers to TRUE and then start from the beginning (Data Pre-processing).


```{r}
outlg <- round(length(sc@fdata[,1])/200) # The cell will be considered as an outlier if it has a minimum of 0.5% of the number of filtered genes as outlier genes.
Outliers <- FindOutliersKM(sc, K=2, outminc=5,outlg=outlg,probthr=.5*1e-3,thr=2**-(1:40),outdistquant=.75, plot=TRUE, quiet=FALSE)
```

#### 2.1.4. Cellular pseudo-time ordering based on k-means clusters

```{r}
sc <- KmeanOrder(sc,quiet=FALSE, export=FALSE)
plotOrderKMtsne(sc)
```

#### 2.1.5 Plotting the K-means clusters in heatmap

The similarities between single-cells were computed by Euclidean distances of Pearson transcriptome correlation matrix. Based on these similarities, a heatmap portrayal of cell-to-cell distances was plotted using Euclidean as the distance measure and the single linkage as the clustering method, cluster centers were ordered by hierarchic clustering.


```{r}
KMclustheatmap(sc,hmethod="single", plot=TRUE)
```

#### 2.1.6 Plotting the gene expression of a particular gene in a tSNE map


```{r}
plotExptSNE(sc, "ENSG00000101346")
```

## 3. Determining differentially expressed genes (DEGs) For K-means Clustering

Differentially expressed genes between individual clusters are identified using the significance analysis of sequencing data (SAMseq), which is a new function in significance analysis of microarrays (Li and Tibshirani 2011) in the samr package v2.0 (Tibshirani et all., 2015). SAMseq is a non-parametric statistical function dependent on Wilcoxon rank statistic that equalizes the sizes of the library by a resampling method accounting for the various sequencing depths. The analysis is implemented over the pure raw dataset that has the unnormalized expression read counts after excluding the ERCCs. Furthermore, DEGs in each cluster comparing to all the remaining clusters are determined using binomial differential expression, which is based on binomial counting statistics.

## 3.1 Identifying DEGs using SAMseq

The user can define DEGs between all clusters generated by either K-means or model based clustering by applying the “DEGanalysis” function. Another alternative is to define DEGs between particular clusters generated by K-means clustering by applying the “DEGanalysis2clust” function. The outcome of these two functions is a list of two components:
- The first component is a data frame showing the Ensembl gage name and the symbole of the detected DEGs
- The second component is table showing the comparisons, Target cluster, Number of genes and the File name. This component will be used for the downstream analysis.

### 3.1.1 Determining DEGs between two particular clusters

```{r degKM}
####### differential expression analysis between cluster 1 and cluster 3 of the Model-Based clustering using FDR of 0.05
cdiff <- DEGanalysis2clust(sc,Clustering="K-means",K=2,fdr=0.05,name="Name",First="CL1",Second="CL2",export=FALSE,quiet=TRUE)
```

```{r}
#### To show the result table
head(cdiff[[1]]) # The first component
head(cdiff[[2]]) # The second component
```

### 3.1.2 Determining DEGs between all clusters


**¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤ Running this cell will overwrite the previous one ¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤¤**


```{r}
cdiff <- DEGanalysis(sc,Clustering="K-means",K=2,fdr=0.05,name="all_clusters",export=FALSE,quiet=TRUE) ####### differential expression analysis between all clusters
```

```{r}
#### To show the result table
head(cdiff[[1]]) # The first component
head(cdiff[[2]]) # The second component
```

#### 3.2 Identifying DEGs using binomial differential expression

The function KMClustDiffGenes identifies differentially regulated genes for each cluster of the K-means clustering in
comparison to the ensemble of all cells. It returns a list with a data.frame element for each cluster that contains the mean expression across all cells not in the cluster (mean.ncl) and in the cluster (mean.cl), the fold-change in the cluster versus all remaining cells (fc), and the p-value for differential expression between all cells in a cluster and all remaining cells. The p-value is computed based on the overlap of negative binomials fitted to the count distributions within the two groups akin to DESeq.


```{r}
cdiffBinomial <- KMClustDiffGenes(sc,K=2,fdr=.01,,quiet=TRUE) ########## Binomial differential expression analysis
#### To show the result table
head(cdiffBinomial[[1]]) # The first component
head(cdiffBinomial[[2]]) # The second component
```

### Plotting the DEGs

Volcano plots are used to readily show the DEGs by plotting significance versus fold-change on the y and x axes, respectively.

```{r}
name <- cdiffBinomial[[2]][2, 4] # Selecting the DEGs table
U <- read.csv(file=paste0(name), head=TRUE, sep=",")
Vplot <- VolcanoPlot(U, value=0.001, name=name, FS=.5, fc=1)
```

In case the output of this function is the error "Error in plot.window(...): need finite 'ylim' values", `adj` should be set to TRUE.

In case the user would like to get the names of the significant DEGs, then FS should be set to a value of 0.3 or higher.

## 4. Identifying biomarkers (decision trees and networking analysis)

There are several methods to identify biomarkers, among them are decision trees and hub detection through networking analysis. The outcome of STRING analysis is stored in tab separated values (TSV) files. These TSV files served as an input to check both the connectivity degree and the betweenness centrality, which reflects the communication flow in the defined PPI networks

Decision trees are one of the most efficient classification techniques in biomarkers discovery. Here we use it to predict the sub-population of a target cell based on transcriptomic data. Two types of decision trees can be performed: classification and regression trees (CART) and J48. The decision tree analysis is implemented over a training dataset, which consisted of the DEGs obtained by either SAMseq or the binomial differential expression. The performance of the generated trees can be evaluated for error estimation by ten-fold cross validation assessment using the "J48DTeval" and "RpartEVAL" functions. The decision tree analysis requires the dataset to be class vectored by applying the “ClassVectoringDT” function.

```{r}
#sigDEG <- cdiff[[1]]             # DEGs gene list from SANseq
sigDEG <- cdiffBinomial[[1]]      # DEGs gene list from Binomial analysis
First="CL1"
Second="CL2"
DATAforDT <- ClassVectoringDT(sc,Clustering="K-means",K=2,First=First,Second=Second,sigDEG)
```

### 4.1. J48 Decision Tree


```{r}
j48dt <- J48DT(DATAforDT)
summary(j48dt)
```

#### 4.1.1. Evaluating the performance of the J48 Decision Tree


```{r}
j48dt <- J48DTeval(DATAforDT,num.folds=10,First=First,Second=Second)
```

### 4.2. RPART Decision Tree


```{r}
rpartDT <- RpartDT(DATAforDT)
```

#### 4.2.1. Evaluating the performance of the RPART Decision Tree


```{r}
rpartEVAL <- RpartEVAL(DATAforDT,num.folds=10,First=First,Second=Second)
```

### 4.3. Networking Analysis

To define protein-protein interactions (PPI) over a list of genes, STRING-api is used. The outcome of STRING analysis was stored in tab separated values (TSV) files. These TSV files served as an input to check both the connectivity degree and the betweenness centrality, which reflects the communication flow in the defined PPI networks.

### 4.3.1 All DEGs


```{r}
DEGs="All DEGs"
FileName=paste0(DEGs)

data <- cdiffBinomial[[1]] [,2]           # DEGs gene list from Binomial analysis (taking only the firat 100 genes)

ppi <- PPI(data,FileName)

networking <- NetAnalysis(ppi)
networking                            ##### In case the Examine response components=200 and an error "linkmat[i, ]" appeared, that means there are no PPI.
```

```{r}
data=networking[1:25,1]              # plotting the network of the top 25 highly connected genes
network <- Networking(data,FileName,plot_width=25, plot_height=10)
```

### 4.3.2 Particular set of DEGs


```{r}
############ Selecting the DEGs' table  ##############
DEGs=cdiffBinomial[[2]][1,4]     # Up-regulated genes in cluster 1 (from the Binomial analysis)
FileName=paste0(DEGs)

data <- read.csv(file=paste0(DEGs),head=TRUE,sep=",")
data <- data[,3]

ppi <- PPI(data,FileName)

networking <- NetAnalysis(ppi)
networking                            ##### In case the Examine response components=200 and an error "linkmat[i, ]" appeared, that means there are no PPI.

network <- Networking(data,FileName)
```

### 4.4 Gene Enrichment Analysis

```{r}
dbs <- listEnrichrDbs()
head(dbs)
```

```{r}
############ Selecting the DEGs' table  ##############
DEGs=cdiffBinomial[[2]][1,4]     # Up-regulated genes in cluster 1 (from the Binomial analysis)

data <- read.csv(file=paste0(DEGs),head=TRUE,sep=",")
data <- as.character(data[,3])

dbs <- c("KEGG_2013","GO_Molecular_Function_2015", "GO_Biological_Process_2015")
enriched <- enrichr(data, dbs)
KEGG_2013 <- enriched[[1]][,c(1,2,3,9)]
GO_Molecular_Function_2015 <- enriched[[2]][,c(1,2,3,9)]
GO_Biological_Process_2015 <- enriched[[3]][,c(1,2,3,9)]

GEA <- rbind(KEGG_2013,GO_Molecular_Function_2015,GO_Biological_Process_2015)
GEA
```
